Figures_MSA-PD

### 2025-10-03 Manuscript_figures.Rmd
Updated files for manuscript figures in R for revision

## Figures 5d,e & EDF13

@@

### Extended Data Figure 13

@@

meta <- read.delim("metadata.tsv", sep = "\t") %>% 
  filter(sample %in% names(con$samples)) %>% 
  tibble::column_to_rownames("sample")

cao$estimateExpressionShiftMagnitudes()
cao$estimateMetadataSeparation(sample.meta = meta, 
                               space = "expression.shifts")

meta %<>% 
  tibble::rownames_to_column("sample")
```

## 3a

```{r, fig.width=18, fig.height=16}
spc <- con$getDatasetPerCell()

@@

for (cc in colnames(meta)[-1]) {
  mpc[[cc]] <- meta[[cc]][match(mpc$sample, meta$sample)]
}

@@

## 3b

```{r, fig.width = 6, fig.height=4}
cao$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP")$data %>% 
  mutate(condition = unname(sample.groups)) %>% 
  ggplot(aes(x, y, shape = condition, col = condition)) +
  geom_point(aes(size = n.cells)) + scale_size_continuous(trans = "log10", 
                                                          range = c(5 * 0.5, 5 * 1.5), name = "Num. cells") +
  theme_bw() +
  scale_color_manual(values = pal.major) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "", y = "", title = "Condition", shape = "Condition", col = "Condition") +
  dotSize(3)
```

```{r, fig.width = 6, fig.height=4}
cao$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$age %>% setNames(rownames(meta)))$data %>% 
  mutate(condition = unname(sample.groups)) %>% 
  ggplot(aes(x, y, shape = condition, col = color)) +
  geom_point(size = 3) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "", y = "", title = "Age", color = "Age (y)", shape = "Condition")
```

```{r, fig.width = 6, fig.height=4}
cao$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)))$data %>% 
  mutate(condition = unname(sample.groups)) %>% 
  ggplot(aes(x, y, shape = condition, col = color)) +
  geom_point(size = 3) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "", y = "", title = "Brain bank", shape = "Condition", color = "Brain bank")
```

```{r, fig.width = 6, fig.height=4}
cao$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)))$data %>% 
  mutate(condition = unname(sample.groups)) %>% 
  ggplot(aes(x, y, shape = condition, col = color)) +
  geom_point(size = 3) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "", y = "", title = "Extract", shape = "Condition", color = "Extract")
```

```{r, fig.width = 6, fig.height=4}
cao$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)))$data %>% 
  mutate(condition = unname(sample.groups)) %>% 
  ggplot(aes(x, y, shape = condition, col = color)) +
  geom_point(size = 3) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = "", y = "", title = "Flowcell", shape = "Condition", color = "Flowcell")
```

@@

@@

@@

# Extended Data Figure 6

@@

## 6a

@@

## 6b

@@

@@

@@

# Extended Data Figures 11 & 12

@@

## 11a

@@

## 11b

@@

## 12

@@

# Extended Data Figure 14

@@

## 15a

@@

## 15b

@@

## 16c

@@

## 17a

@@

## 17b

@@

data.frame(cell = dat.scores$X %>% gsub("one_", "!!", .),
           score = dat.scores$norm_score,
           type = anno.major[match(dat.scores$X,
                                   names(anno.major))]) %>%
  filter(cell %in% names(anno.micro)) %>% 
  mutate(type = factor(anno.micro[.$cell])) %>% 
  group_by(type) %>%
  summarize(m = mean(score)) %>%
  filter(!is.na(type)) %>% 
  mutate(type = as.character(type)) %>% 
  arrange(desc(m)) %>% 
  mutate(type = factor(type, levels = type), sign = dat.sign$sign[match(type %>% levels, dat.sign$type)]) %>%
  mutate(type = factor(type, levels = type)) %>% 
  ggplot(aes(type, m, fill = type)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = sign), vjust = -0.5, nudge_y = -0.05) +
  theme_bw() + 
  theme(line = element_blank()) +
  labs(y = "Mean score", x = "", title = "scDRS for microglia") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ylim(c(0, 2)) +
  scale_fill_manual(values = pal.major)
```

@@

>>>>

## Figures 5d,e & EDF14

@@

### Extended Data Figure 14

@@

meta <- read.delim("metadata.tsv", sep = "\t") %>%
  filter(sample %in% names(con$samples)) %>%
  tibble::column_to_rownames("sample") %>% 
  dplyr::select(-condition)

meta %<>%
  mutate(sample = rownames(.))

spc <- con$getDatasetPerCell()

@@

for (cc in colnames(meta)) {
  mpc[[cc]] <- meta[[cc]][match(mpc$sample, meta$sample)]
}

@@

@@

## 4a

```{r}
con_pd <- con$clone()
con_pd$embedding <- con$embeddings$UMAP

con_pd$samples <- con$samples %>% .[!grepl("MSA", names(.))]

sample.groups <- con_pd$samples %>% 
  names() %>% 
  setNames(grepl.replace(., c("CTRL","PD")), .)

cao_pd <- Cacoa$new(data.object = con_pd, 
                 sample.groups = ifelse(grepl("CTRL", sample.groups), "CTRL", "PD") %>% 
                   setNames(names(sample.groups)), 
                 cell.groups = anno.major, 
                 ref.level = "CTRL", 
                 target.level = "PD", 
                 n.cores = 32)

cao_pd$estimateExpressionShiftMagnitudes()
cao_pd$estimateMetadataSeparation(sample.meta = meta %>% filter(!grepl("MSA", sample)) %>% dplyr::select(-disease_duration, -subtype, -sample), 
                               space = "expression.shifts")
```

```{r, fig.width=20, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_pd$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_pd$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_pd$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 4) -> p

## 4b

```{r}
con_msa <- con$clone()
con_msa$embedding <- con$embeddings$UMAP

con_msa$samples <- con$samples %>% .[!grepl("PD", names(.))]

sample.groups <- con_msa$samples %>% 
  names() %>% 
  setNames(grepl.replace(., c("CTRL","MSA")), .)

cao_msa <- Cacoa$new(data.object = con_msa, 
                 sample.groups = ifelse(grepl("CTRL", sample.groups), "CTRL", "MSA") %>% 
                   setNames(names(sample.groups)), 
                 cell.groups = anno.major, 
                 ref.level = "CTRL", 
                 target.level = "MSA", 
                 n.cores = 32)

cao_msa$estimateExpressionShiftMagnitudes()
cao_msa$estimateMetadataSeparation(sample.meta = meta %>% filter(!grepl("PD", sample)) %>% dplyr::select(-brain_bank, -sample), 
                               space = "expression.shifts")
```

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_msa$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
    cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_msa$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_msa$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

## 4c

```{r}
con_dis <- con$clone()
con_dis$embedding <- con$embeddings$UMAP

con_dis$samples <- con$samples %>% .[!grepl("CTRL", names(.))]

sample.groups <- con_dis$samples %>% 
  names() %>% 
  setNames(grepl.replace(., c("PD","MSA")), .)

cao_dis <- Cacoa$new(data.object = con_dis, 
                 sample.groups = ifelse(grepl("PD", sample.groups), "PD", "MSA") %>% 
                   setNames(names(sample.groups)), 
                 cell.groups = anno.major, 
                 ref.level = "PD", 
                 target.level = "MSA", 
                 n.cores = 32)

cao_dis$estimateExpressionShiftMagnitudes()
cao_dis$estimateMetadataSeparation(sample.meta = meta %>% filter(!grepl("CTRL", sample)) %>% dplyr::select(-sample), 
                               space = "expression.shifts")
```

```{r, fig.width=25, fig.height=8}
cowplot::plot_grid(plotlist = list(
  cao_dis$plotSampleDistances(space = "expression.shifts", show.sample.size = T, method = "UMAP", title = "Condition"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$age %>% setNames(rownames(meta)), title = "Age"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$brain_bank %>% setNames(rownames(meta)), show.sample.size = T, title = "Brain bank"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$disease_duration %>% setNames(rownames(meta)), show.sample.size = T, title = "Disease duration"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$extract %>% setNames(rownames(meta)), show.sample.size = T, title = "Extract batch"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$flowcell %>% setNames(rownames(meta)), show.sample.size = T, title = "Flowcell"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", show.sample.size = T, sample.colors = meta$pmi %>% setNames(rownames(meta)), title = "PMI"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$sex %>% setNames(rownames(meta)), show.sample.size = T, title = "Sex"),
  cao_dis$plotSampleDistances(space = "expression.shifts", method = "UMAP", sample.colors = meta$subtype %>% setNames(rownames(meta)), show.sample.size = T, title = "Subtype"),
  cao_dis$test.results$metadata.separation$pvalues %>%
    data.frame(p = .) %>%
    mutate(pp = -log10(p)) %>%
    tibble::rownames_to_column("covariate") %>%
    ggplot(aes(covariate, pp)) +
    geom_bar(stat = "identity", fill = "lightblue4") +
    ylim(0, 2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    labs(x = "", y = "-log10(adj. p)", title = "Association significance") +
    geom_hline(yintercept = -log10(0.05), colour = "red3")
), ncol = 5) -> p

# Extended Data Figure 5

@@

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "flowcell", "extract"), # Omitting brain bank
                                     show_donor_ids = TRUE,
                                     add_meta_associations="pval")

@@

@@

## 7a

@@

## 7b

@@

# Extended Data Figure 9

@@

These plots were created with results from LIANA in Python.

# Extended Data Figure 11

@@

```{r, fig.width=8, fig.height=4}
res.old <- c("TSPO", "COQ2", "MAPT", "FBXO47", "ELOVL7", "EDN1", "GAB1", "TENM2", "RABGEF1", "PLA2G4C", "INPP4B", "ZIC1", "ZIC2", "ZIC3", "ZIC4", "SNCA", "TPPP") %>% 
  lapply(\(gene) {
    idx <- cm.pseudo %>% 
    colnames() %>% 
    data.frame(id = .) %>% 
    mutate(condition = strsplit(id, "_|!!") %>% sget(1),
           ct = strsplit(id, "!!") %>% sget(2)) %>% 
    mutate(ord = order(condition, ct))

  stat.test <- plot.dat %>% 
    group_by(anno) %>% 
    rstatix::wilcox_test(value ~ condition)

  stat.omnibus <- plot.dat %>% 
    group_by(anno) %>% 
    rstatix::kruskal_test(value ~ condition)

  out <- data.frame(Gene = gene, celltype = stat.test$anno, omnibus.old = stat.omnibus$p, contrast = sapply(seq(nrow(stat.test)), \(x) paste(stat.test$group1[x], stat.test$group2[x], sep = " - ")), padj.old = stat.test$p.adj)

  return(out)
  }) %>% 
  bind_rows()
```

# Extended Data Figures 12 & 13

@@

## 12a

@@

## 12b

@@

## 13

@@

@@

## 16a

@@

## 16b

@@

@@

## 18a

@@

## 18b

@@

p <- data.frame(cell = dat.scores$X %>% gsub("one_", "!!", .),
           score = dat.scores$norm_score,
           type = anno.major[match(dat.scores$X,
                                   names(anno.major))]) %>%
  filter(cell %in% names(anno.micro)) %>% 
  mutate(type = factor(anno.micro[.$cell])) %>% 
  group_by(type) %>%
  summarize(m = mean(score)) %>%
  filter(!is.na(type)) %>% 
  mutate(type = as.character(type)) %>% 
  arrange(desc(m)) %>% 
  mutate(type = factor(type, levels = type), sign = dat.sign$sign[match(type %>% levels, dat.sign$type)]) %>%
  mutate(type = factor(type, levels = type)) %>% 
  ggplot(aes(type, m, fill = type)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = sign), vjust = -0.5, nudge_y = -0.05) +
  theme_bw() + 
  theme(line = element_blank()) +
  labs(y = "Mean score", x = "", title = "scDRS for microglia") +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ylim(c(0, 2)) +
  scale_fill_manual(values = pal.major)

## 18c

@@

# Extended Data Figure 19
